---
title: "Introduction to package `isismdl`"
author: "Rob van Harrevelt"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
linkcolor: blue    
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  
---

# Introduction

This introductiomn shows how package `isismdl` can be
used to solve a simple example mode, 
a dynamical version of the ISLM model.

[Here follows a description of the ISLM model]

<!--   
PREPARATIONS
-->
```{r echo = FALSE, message = FALSE}

library(isismdl)

# input/output chunks are separated
# R inputlines start with >, outputlines with nothing
knitr::opts_chunk$set(collapse = FALSE, comment = "", prompt = TRUE)
```

## The model file

The model file should be defined on an external ASCII file. A detailed
discussion about the syntax of this file is provided in the Reference
Manual of Isis.

For the example ISLM-model, the model files has the following contents

```{r echo = FALSE, comment=''}
cat(readLines('islm.mdl'), sep = '\n')
```

## Compiling the model

The function `compile_mdl` compiles the model and 
generates internal equation code that is used to evaluate the
model equations:
```{r}
compile_mdl("islm.mdl")
```

Two files are generated:

* `islm.mif` The model information file (mif-file). This is a binary file
contains the interal code for the evaluation of the model
equations.
* `islm.mrf` The model reference file (mrf file), a text file with information about
the structure of the model.

If the the compiler detects an error in the model file, an additional
file `islm.err` is created. This file contains a list of errors.

The mrf file for our ISLM looks like this:

```{r echo = FALSE}
cat(readLines('islm.mrf'), sep = '\n')
```

The mrf file contains tehnical information about the model. For example,
it shows the feedbackvariables of the model. It is important to know
the feedback variables of the model, because for these variables initial values
should be provided when solving the model.

## `IsisMdl` objects

The function `read_mdl` reads the mif files and creates
a `Isismdl` object.

```{r}
mdl <- read_mdl("islm.mif")
mdl
```

An `IsisMdl` object is an R6 class object. R6 
classes behave quite differently than the more familiar S3 and S4 classes. 
For example, for R6 classes methods are part of the object itself and not of 
generic functions. R6 classes behave simular to classes in
object oriented languages such as Java, C++ or Python. 

For example, the method `get_params()` can be used to
obtain the values of the model parameters. 
For example, to obtains the values of parameters `m0` and `c1`, use
```{r}
mdl$get_param(names = c("m0", "c1"))
```

Methods starting with `get_`, are ofter called a "getter" methods.
There are also corresponding "setter" methods. For example
```{r}
mdl$set_param(list(m0 = 100))
mdl$get_param("m0")
```
Input for method `set_params()` is a named list.

## Cloning `IsisMdl` objects

Consider the following assignment:
```{r}
mdl2 <- mdl
```
Now variables `mdl2` and `mdl` refer to the same object. If you modify
`mdl2`, then also `mdl` is modified.
```{r}
mdl2$set_param(list(m0 = 75))
mdl$get_param("m0")
```
The usual copy-on-modify semantics that it used for convential
R objects such as S3 or S4 classes does not apply to R6 classes.

If you want to create a copy of the model, use the `clone()` method with
argument `deep = TRUE` :
```{r}
mdl2 <- mdl$clone(deep = TRUE)
mdl2$set_param(list(m0 = -9999))
mdl$get_param("m0")
```

## Solving the model \label{sec:solving}

Suppose that we want to solve the model for the period from 2016 to
2021. Then we need the values of the exogenous variables for that period.
For the feedback variables `y` and `r` we also need initial
starting values for the solution period.
Since `r`, `y` and `y_d` have a lag 1, we also need values for these
variables in 2015. 

To supply the required data to the model, we first a `regts` 
timeseries object. For the exogenous variables `g` and `ms` we
assume an annual growth of 1.5% after 2016.
```{r}
# exogenous variables:
g  <- regts(210 * cumprod(rep(1.015, 6)), start = "2016")
ms <- regts(200 * cumprod(rep(1.015, 6)), start = "2016")
# feedback variables (with lag):
r <- regts(3.4, period = "2015/2021")
y <- regts(980, period = "2015/2021")
# lagged variable
yd <- regts(790, start = "2015")
data <- cbind(g, ms, r, y, yd)
data
```

To transfer these variables to the model, we first
have to define the model period. For this we use the
metbhod `set_period()`:
```{r}
mdl$set_period(regperiod_range("2016", "2021"))
mdl
```
The data period is the model period extended with the lag and lead period.
You can also retrieve the model period and model data period 
with the methods `get_period()` and `get_data_period`.

Now can transfer the values in `regts` object data the model
with method `set_data()`:
```{r}
mdl$set_data(data)
```

Method `set_data()` is described in more detail in section \ref{sec:data}.
To check the model data, we use
```{r}
mdl$get_data()
```
Model variables that we have not explicitly set all have value `NA`. 
Values for these variables will be computed when solving the model.

We are now ready to solve the model:
```{r}
mdl$solve()
```

The method  `get_data()` can be used to retrieve the solution.
```{r}
mdl$get_data()
```

## Transferring data to the model \label{sec:data}

In section \ref{sec:solve}, we used method `set_data` to initialize
the model variables. This section gives a a more detailed  description of this 
method.

The method compares the column names of the input and 
updates the corresponding model variable. However, sometimes
the input does not have column names or not the column names 
that you need. Univariate timeseries typically have no column
names (see the vignette "Introduction to package regts" for
more information about this issue). In that case, we use argument
`names` to supply the timeseries namnes.

The methods `set_ca`, `set_fix`, and `set_fit` can be used
to set constant adjustments, fix values and fit targets. 
They work similar to `set_data`. 
For example, suppose that we have new information about exogenous 
variable `g` for periods 2016 and 2017:
```{r}
g <- regts(c(213, 218),  period = "2016/2017")
```
Since `g` does not have a column name, we have to specify the `names`
argument:
```{r}
mdl$set_data(g, names = "g")
mdl$get_data("g")
```

This example also shows that the model variables are only
modified for the intersection of the period range of the
input timeseries and the model data period.

## Fixing variables

There are to type of equations in Isis models:
the so called `frml` (formula) equations, also
called behavioural equations, and identities.  The frml equations


Endogenous variables that occur in the left hand side of frml equations
